---
title: "temp"
author: "Jasen Zhang"
date: "5/16/2021"
output: html_document
---

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(lme4)
library(nlme)
library(caret)
library(corpcor)
library(Hmisc)
```

# Functions

```{r}
replace_period_with_hyphen <- function(arr){
  gene_pat2 <- c()
  
  for(i in arr){
    temp1 <- str_replace_all(i, '\\.', '-')
    gene_pat2 <- append(gene_pat2, temp1)

  }
  gene_pat2
}

sample_to_patient <- function(arr){
  gene_pat2 <- c()
  
  for(i in arr){
    temp1 <- substr(i, 1, 12)
    gene_pat2 <- append(gene_pat2, temp1)

  }
  gene_pat2
}

#------------------------------------
# input: 
#   df: n x p matrix of observations
#   n: n
#   names: the names of the subjects 
#   
# output:
#   K: n x n correlation matrix
#------------------------------------

get_pearson_correlation_matrix <- function(df, n, names){
  K <- matrix(rep(0,n^2), n, n)
  
  for(i in 1:n){
    for(j in 1:i){
      temp1 <- as.numeric(unname(df[i,]))
      temp2 <- as.numeric(unname(df[j,]))
      temp_correlation <- cor(temp1, temp2)
      K[i,j] <- temp_correlation
      K[j,i] <- temp_correlation
    }
    print(i)
  }
  
  rownames(K) <- names
  colnames(K) <- names
  K
}

#-----------------------------------------------------------------------
# Goal: Converts the factors of a dataframe to 1/0 columns
# input: 
#   df: dataframe of covariates
#   name: vector of column names of factor variables
#   
# output:
#   df: new dataframe of covariates with factors split into 1/0 columns
#------------------------------------------------------------------------

split_factor <- function(df, names){
  for(name in names){
    temp <- dplyr::pull(df, name)
    lev <- unique(temp)
    lev <- lev[1:length(lev) - 1]
    for(i in lev){
      i2 <- paste(name, i, sep = '.')
      df <- df  %>% mutate(!!as.name(i2) := as.numeric(temp == i))
    }
    df <- df %>% select(- !!as.name(name))
  }
  df
}
#--------------------------------------------------------------
# Goal: implementation of GLS theoretical solution
# input:
#   X: n x p covariate matrix
#   K: n x n error matrix
#   Y: n x 1 response matrix
#   n: number of observations
# 
# output:
#   a list containing the following:
#     1: beta_hats
#     2: standard errors of the beta_hats
#     3: estimate of the variance multiplied by K
#     4: y_hats
#     5: residuals
#     6: rmse
#-------------------------------------------------------------------
    
GLS_solver <- function(X,K,Y,n){
  beta <- pseudoinverse(t(X) %*% pseudoinverse(K) %*% X) %*% t(X) %*% pseudoinverse(K) %*% Y
  var_mle <- as.numeric((1/n) * t(Y - X %*% beta) %*% pseudoinverse(K) %*% (Y - X %*% beta))
  beta_var <- var_mle * pseudoinverse(t(X) %*% pseudoinverse(K) %*% X)
  
  beta_se <- as.matrix(sqrt(diag(beta_var)))
  
  y_hat <- X %*% beta
  e <- Y - y_hat
  rmse <- sqrt(mean(e^2))
  
  answer <- list(beta, beta_se, var_mle, y_hat, e, rmse)
  answer
}


#--------------------------------------------------------------------
# goal: create a vector that partitions n objects into k folds
# input:
#   n: how many observations
#   num_folds: how many folds
#   
# output:
#   allocation2: vector describing how many objects in each fold
#-------------------------------------------------------------------


create_fold_allocation <- function(n, num_folds){
  remainder <- n %% num_folds
  quotient <- floor(n / num_folds)
  
  allocation <- rep(quotient, num_folds)
  
  remainder_allocation <- c( rep(1, remainder), rep(0, num_folds - remainder))
  allocation2 <- allocation + remainder_allocation  
  
  allocation2
}


#----------------
# goal: to calculate the rmse of 10 fold cross validation
# input: 
#   df_train: the training set dataframe, only used for column names
#   K: the training set covariance error matrix
#   X: the training set covariates
#   Y: the training set response variable
#   
# output:
#   rmse: the overall root mean square error of the cross validation
#----------------

GLS_CV <- function(df_train, K, X, Y, seed){
  doomed <- F
  IDs <- rownames(df_train)
  set.seed(seed)
  random_IDs <- sample(IDs, length(IDs), replace = F)
  
  allocation <- create_fold_allocation(nrow(df_train), 10)
  
  fold_ID <- partition.vector(random_IDs, allocation)
  
  fit_cv <- data.frame()
  
  for(i in 1:10){
    df_validation <- df_train[fold_ID[[i]],]
    
    remaining_IDs <- fold_ID
    remaining_IDs[[i]] <- NULL
    
    remaining_IDs <- unname(unlist(remaining_IDs))
    
    df_train_2 <- df_train[remaining_IDs,]
    
    K_CV <- K[remaining_IDs, remaining_IDs]
    
    X_CV <- X[remaining_IDs,]
    
    X_validation <- X[fold_ID[[i]],]
      
    Y_CV <- Y[remaining_IDs,]
    n_CV <- nrow(df_train_2)
    
    gls_model_CV <- try(GLS_solver(X_CV,K_CV,Y_CV,n_CV))
    if(length(gls_model_CV[[1]]) == 1){
      doomed <- T
      break
    }
    
    betas <- gls_model_CV[[1]]
    
    y_pred <- X_validation %*% betas
    
    
    fit_cv <- rbind(fit_cv, y_pred)
    
  }
  if(doomed){
    rmse <- NA
  }
  else{
    fit_cv2 <- unlist(fit_cv)
    Y_reshuffled <- Y[random_IDs,]
    
    rmse <- sqrt(mean((Y_reshuffled - fit_cv2)^2))
  }

  rmse
}

```

# Read the datasets

```{r}
genes <- read.csv('TCGA-LUSC.htseq_counts.tsv', sep = '\t')
phenotype <- read.csv('TCGA-LUSC.GDC_phenotype.tsv', sep = '\t')
ensembl_id <- genes$Ensembl_ID
```

# Get the sample ID's from all 2 dataframes

```{r}
genes_id <- colnames(genes)
genes_id2 <- replace_period_with_hyphen(genes_id)

phenotype_id <- phenotype$submitter_id.samples
```


# Figure out which samples are in all 3 dataframes

```{r}
all_id <- intersect(genes_id2, phenotype_id)
all_patients <- sample_to_patient(all_id)
```

# Filter the datasets by common samples

```{r}
genes_index <- genes_id2 %in% all_id
phenotype_index <- phenotype_id %in% all_id

phenotype2 <- phenotype[phenotype_index,]
genes2 <- as.data.frame(t(genes[, genes_index]))
colnames(genes2) <- ensembl_id
rownames(genes2) <- replace_period_with_hyphen(rownames(genes2))
```


# Getting annotation file - skip if you already have annotations.csv

```{r}
zz <- gzfile('Homo_sapiens.GRCh38.104.gtf.gz','rt')  
dat <- read.csv(zz, skip = 5, header = F, , sep = '\t')
```

# Parsing annotation file - skip if you already have annotations.csv


```{r}
df <- dat[,9]

results_df <- data.frame()

for(i in 1:length(df)){
  slice <- df[i]
  
  slice2 <- str_replace_all(slice, ';', '')

  slice3 <- unlist(str_split(slice2, ' '))
  
  gene_id_index <- match('gene_id', slice3)
  gene_id <- slice3[gene_id_index + 1]
  
  gene_biotype_index <- match('gene_biotype', slice3)
  gene_biotype <- slice3[gene_biotype_index + 1]
  
  result <- c(gene_id, gene_biotype)
  
  if(! gene_id %in% results_df$X.ENSG00000284662.){
    results_df <- rbind(results_df, result)
  }
  
  
}

colnames(results_df) <- c('gene_id', 'annotation')

write.csv(results_df,"annotations.csv")

```

# Getting annotation file

```{r}
annotations <- read.csv('annotations.csv') %>% select(-X)



ensembl_id2 <- c()
annotation_2 <- c()

for(i in ensembl_id){
  i2 <- substr(i, 1, 15)
  ensembl_id2 <- append(ensembl_id2, i2)
  
  if(i2 %in% annotations$gene_id){
    the_index <- which(annotations$gene_id == i2)
    the_type <- annotations[the_index,2]
    annotation_2 <- append(annotation_2, the_type)
  }
  else{
    annotation_2 <- append(annotation_2, NA)
  }
}
```


# Filtering only protein coding genes

```{r}
pc_index <- annotation_2 == 'protein_coding'
pc_index2 <- which(pc_index %in% c(TRUE))
genes_pc <- genes2[, pc_index2]
```

```{r}
genes_var <- genes_pc %>%
  select(-(names(genes_pc[, sapply(genes_pc, function(v) var(v, na.rm=T)==0)])))

```


# Getting first n PC's of the genes

```{r}
pca1 <- prcomp(genes_var)

top_100_pc <- as.data.frame(pca1$x[,1:100])
```

# Interpreting the principal components

```{r}
important_genes <- data.frame()

evecs <- pca1$rotation

evecs1 <- evecs[,1:11]

for(i in 1:11){
  slice <- evecs1[,i]
  top_5 <- order(abs(slice))[1:5]
  top_5_genes <- names(slice)[top_5]
  important_genes <- rbind(important_genes, top_5_genes)
}

```



# Getting relevant phenotype covariates

```{r}
covariates <- c('gender.demographic',
                'race.demographic',
                'ethnicity.demographic',
                'age_at_diagnosis.diagnoses',
                'pack_years_smoked.exposures',
                'tobacco_smoking_history')


phenotype3 <- phenotype2 %>% select(one_of(covariates))

# Filling in NA's of pack years smoked

phenotype3$pack_years_smoked.exposures[is.na(phenotype3$pack_years_smoked.exposures)] <- 0

# Changing age at diagnosis in years

phenotype3$age_at_diagnosis.diagnoses <- phenotype3$age_at_diagnosis.diagnoses/365.25

rownames(phenotype3) <- phenotype2$submitter_id.samples
phenotype3$sample_ID <- phenotype2$submitter_id.samples
```


# Merging the dataframes

```{r}
phenotype3$ID <- rownames(phenotype3)
top_100_pc$ID <- rownames(top_100_pc)



df_model <- phenotype3 %>% full_join(top_100_pc, by = 'ID')

df_model2 <- na.omit(df_model)
df_model2$ID <- 1:nrow(df_model2)
```



# fixing the indices of top_100_pc

```{r}
non_na_ID <- as.numeric(rownames(df_model2))
genes_pc2 <- genes_pc[non_na_ID,]

top_100_pc <- top_100_pc[non_na_ID,]
top_100_pc <- top_100_pc %>% select(-ID)
```

#creating and saving train and test df

```{r}
n <- nrow(df_model2)

set.seed(284)
test_ID <- sample(rownames(df_model2), 0.2*n, replace = F)
train_ID <- setdiff(rownames(df_model2), test_ID)


df_test <- df_model2[test_ID,]
df_train <- df_model2[train_ID,]

save(df_test, file = "df_test_v2.RData")
save(df_train, file = "df_train_v2.RData")
```


